---
title: "Analysis"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    theme: journal
    fig_width: 6
  pdf_document:
    toc: yes
  html_notebook: 
    toc: yes
    theme: flatly
    toc_depth: 5
    highlight: pygments
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6, warning = FALSE
)
```

```{r packages, include=FALSE}
require(caret)
require(dplyr)
require(FMradio)
require(rags2ridges)
require(rpart)
require(randomForest)
require(ROCR)
```

## Data loading

```{r data loading, echo=FALSE}
# Invoke data
load('ADdata.Rda')
```

## Differential expression of metabolites per ApoE genotype

### Data Preparation

```{r Y}
# Transpose, standardize and store the metabolite data in Y
Y <- as.matrix(ADmets)
```

```{r df}
# Store the ApoE genotype in x as a factor
x <- as.factor(geno$APOEb)
sex <- as.factor(geno$sex)
df <- cbind.data.frame(Y,x)
```

### Analysis of Variance (ANOVA)

```{r aov}
# Store the ANOVA model summaries in summaries
summaries <- purrr::map(df[,1:230], ~summary(aov(.x ~ df$x)))
```

#### Correction for Multiple Testing

```{r FDR adjust}
##Extract and adjust p-values for FDR control
#Create a list to store the p-values
p_values <- list(1:230)

#Extract the p-values of the F-tests from the aov summaries list and store them in p_values
for (i in 1:230){
  p_values[[i]] <- summaries[[i]][[1]][["Pr(>F)"]][[1]]
}

# Coerce p_values to dataframe and transpose it 
p_values <- t(data.frame(p_values))

# Set as row names the metabolites
p_values <- data.frame(p_values, row.names = colnames(Y))

# Calculate the FDR-adjusted p-values
p_values$p_adj <- p.adjust(p_values[,], method = "fdr")
```


```{r}
# Filter out the non-significant (a=0.05) FDR-adjusted p-values
filter(p_values, p_adj < 0.05)
```

## Classification of metabolites on ApoE class

A function was created to streamline the analysis of multiple (tidy tabular standardised) predictors and response. The function *fit* fits a Logistic Regression, a Decision Tree and a (bagging) Random Forest. It takes the arguments:

-   X = data frame of predictors,

-   Y = response vector,

-   max_depth = maximum depth of the Decision Tree,

-   mtry = the number of random predictors to be used per training of the Random Forest and

-   seed

It first binds *X* and *Y* into *df* which is then split into a training and test set (maintaining class balances). Next it fits the 3 models, returns a dataframe with their repeated (100 times) 10-fold Cross-Validation(CV) performance metrics and plots their ROC curves.

```{r fit function def}
  ## Function fit
fit <- function(title,
                X,
                y,
                split = 0.8,
                model,
                times,
                ctrl= trainControl(method = "cv",
                                   number = 5,
                                   classProbs = TRUE,
                                   summaryFunction = twoClassSummary),
                grid = NULL,
                seed=123,
                auc = numeric(times),
                metrics  = matrix(nrow = 11, ncol = times)) {
    ### Training-test split
  #Merge X and y into df
  df <- cbind.data.frame(X, y)
  
  # Perform 10-fold CV repeated 100 times
  for (i in 1:times) {
    set.seed(seed + i)  # Set a different seed for each iteration
    # Create a matrix with the training indexes
    split_index <- createDataPartition(df$y, p = split, list = FALSE)
    
    # Create the training set
    train <- df[split_index,]
    
    # Create the test set
    test <- df[-split_index,]
    
    # Store the test y in labels
    labels <- test$y
    labels.n <- as.numeric(labels) - 1
    
    # Train the model
    mdl <- train(y ~ .,
      train,
      method = model,
      tuneGrid = grid,
      metric = "ROC",
      trControl = ctrl
    )
    
    # Create a confusion matrix and get performance metrics from caret
    cm <- confusionMatrix(labels, predict(mdl, test), positive = 'E4YES')
    # Store the model performance metrics in the cumulative data frame
    metrics[,i] <- cm$byClass
    # Store cumulative ROC curve
    yhat <- predict(mdl, test, type= 'prob')
    auc[i] <- pROC::auc(labels.n,
                         yhat[,2],
                         levels=c(0,1),
                         ci=TRUE,
                         boot.n=100,
                         ci.alpha=0.95,
                         stratified=T,
                         quiet=T)
  } 
  metrics <- data.frame(c(rowMeans(metrics, na.rm=TRUE), mean(unlist(auc))),
                             row.names = c(names(cm$byClass),'AUC'))
  names(metrics) <- title
  return(metrics)
}
```

### Original data

```{r fit_original}
X <- Y
#Invoke the function
## Logistic Regression
metrics.lr <- fit(title='Logistic Regression',
                  df=df,
                  model='glm',
                  times=100,
                  grid=NULL)

## Decision Tree
metrics.tree <- fit(title='Decision Tree',
                  df=df,
                  model='rpart2',
                  grid = expand.grid(maxdepth=5),
                  times=100)

## Ensemble Classifiers
metrics.rf <- fit(title='Random Forest',
                  df=df,
                  model='rf',
                  grid= expand.grid(mtry=10),
                  times=100)

metrics.gbm <- fit(title='Stochastic Gradient Boosting',
                   df=df,
                   model='gbm',
                   grid = expand.grid(gbm$bestTune),
                   times=10)
metrics <- cbind(metrics.lr, metrics.tree, metrics.rf)
metrics
```

### Using 6 ML-estimated factor scores

```{r filterX, include=FALSE}
set.seed(123)
cov <- cor(X)

#Find redundant features
filter <- RF(cov)

#Filter out redundant features
filtered <- subSet(X, filter)

#Regularized correlation matrix estimation
M <- regcor(filtered)
```

```{r optCor, echo=FALSE, include=FALSE}
#Get the regularized correlation matrix of the filtered dataset
R <- M$optCor

#Get the Guttman bounds for R
Guttman.bounds <- dimGB(R)
```

```{r dimVar, eval=FALSE, include=FALSE}
#Assess the proportion of cumulative variances for 6 factor solutions
dimVAR(R, maxdim=6)
```

```{r mlfa, include=FALSE}
mlfa <- mlFA(R, m=6)
thomson <- facScore(filtered,mlfa$Loadings,mlfa$Uniqueness)
```

```{r fit_thomson, warning=FALSE}
fit(X = thomson,
    Y = sampleInfo,
    max_depth = 3,
    mtry = 5)
save(metrics, file='metrics_fa.Rdata')
```

## Network Analysis

Code from rags2ridges

```{r preparation}
# Store all observations of Class 1 in C1
C1 <- scale(t(ADmetabolites[, sampleInfo$ApoEClass == "Class 1"]))
# Store all observations of Class 2 in C2
C2 <- scale(t(ADmetabolites[, sampleInfo$ApoEClass == "Class 2"]))

#Get the covariance matrices of C1 and C2
S1<- covML(C1)
S2<- covML(C2)

#Store them in a list
S<- list(S1 = S1, S2 = S2)

#Get the total number of samples
n <- c(nrow(S1), nrow(S2))

# Create a list of fused covariance matrices T
Ts<- default.target.fused(Slist = S, ns = n, type = "DUPV")

```

```{r, eval=FALSE}
# Get the optimal lambdas per class and fused with 10-fold CV
set.seed(8910)
optf <- optPenalty.fused(Ylist = Ys,
                         Tlist = Ts,
                         lambda = as.matrix(cbind(c("ridge1", "fusion"),
                                                  c("fusion", "ridge2"))),
                         cv.method = "kCV",
                         k = 10,
                         verbose = FALSE)
```

```{r sparsify}
# Create a list of the two-class data Y
Ys<- list(C1 = C1, C2 = C2)

load('opt.rda')
Ps <- optf$Plist
# Get the sparsified high precision matrices, correcting FDR at .001
P0s <- sparsify.fused(Ps,
                      threshold = "localFDR",
                      FDRcut = 0.999,
                      verbose = FALSE)
```

```{r GGMs per class and diff, message=FALSE}
#Merge the sparse high precision matrices
TST <- Union(P0s$C1$sparseParCor, P0s$C2$sparseParCor)
PCclass1 <- TST$M1subset
PCclass2 <- TST$M2subset

# Create a color map per metabolite class
Colors <- rownames(PCclass2)
Colors[grep("Amine", rownames(PCclass2))]     <- "lightblue"
Colors[grep("Org.Acid", rownames(PCclass2))]  <- "orange"
Colors[grep("Lip", rownames(PCclass2))]       <- "yellow"
Colors[grep("Ox.Stress", rownames(PCclass2))] <- "purple"


set.seed(111213)
opar <- par(mfrow = c(1, 3))
#Plot the sparsified ridge matrix of AD Class 2
Coords <- Ugraph(PCclass2, type = "fancy", lay = "layout_with_fr",
    Vcolor = Colors, prune = FALSE, Vsize = 7, Vcex = 0.3,
    main = "AD Class 2")
#Plot he sparsified ridge matrix of AD Class 1
Ugraph(PCclass1, type = "fancy", lay = NULL, coords = Coords,
    Vcolor = Colors, prune = FALSE, Vsize = 7, Vcex = 0.3,
    main = "AD Class 1")

#Plot the differential network
DiffGraph(PCclass1, PCclass2, lay = NULL, coords = Coords,
    Vcolor = Colors, Vsize = 7, Vcex = 0.3,
    main = "Differential Network")
par(opar)
```

```{r centrality}
PC0list      <- list(PCclass1 = PCclass1, PCclass2 = PCclass2)
# Get the network statistics
NetStats <- GGMnetworkStats.fused(PC0list)

# Get the centrality degree scores for each class
DegreesAD1   <- data.frame(rownames(NetStats),NetStats$PCclass1.degree)
DegreesAD2   <- data.frame(rownames(NetStats),NetStats$PCclass2.degree)

# Order and show the centrality degree scores
DegreesAD1o  <- DegreesAD1[order(DegreesAD1[, 2], decreasing = TRUE), ]
DegreesAD2o  <- DegreesAD2[order(DegreesAD2[, 2], decreasing = TRUE), ]
head(DegreesAD1o, 7)
head(DegreesAD2o, 7)
```

```{r communities}
# Get the communities per class
set.seed(141516)
opar <- par(mfrow = c(1, 2))
CommC1 <- Communities(PCclass1, Vcolor = Colors, Vsize = 7,  Vcex = 0.3, main = "Modules AD Class 1")

CommC2 <- Communities(PCclass2, Vcolor = Colors, Vsize = 7,  Vcex = 0.3, main = "Modules AD Class 2")
par(opar)
```

```{r degree density plot }
# Plot the densities of centrality degree scores
plot(density(DegreesAD1[, 2]), col = "blue", xlim = c(-1, 8), xlab = "Degree", main = "")
lines(density(DegreesAD2[, 2]), col = "red")
legenda <- c("AD class 1", "AD class 2")
legend(5, 0.5, legend = legenda, lwd = rep(1, 2), lty = rep(1, 2), col = c("blue", "red"), cex = 0.7)
```

```{r wilcoxon rank sum }
#Perform a Wilcoxon signed rank test to test if the centrality degree scores are different between the classes
wilcox.test(DegreesAD1[, 2], DegreesAD2[, 2], paired = TRUE, alternative = "less")
```
