---
title: "RQii - Classification of metabolites on ApoE class"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Introduction

The Tufte handout style is a style that Edward Tufte uses in his books and handouts. Tufte's style is known for its extensive use of sidenotes, tight integration of graphics with text, and well-set typography. This style has been implemented in LaTeX and HTML/CSS^[See Github repositories [tufte-latex](https://github.com/tufte-latex/tufte-latex) and [tufte-css](https://github.com/edwardtufte/tufte-css)], respectively. We have ported both implementations into the [**tufte** package](https://github.com/rstudio/tufte). If you want LaTeX/PDF output, you may use the `tufte_handout` format for handouts, and `tufte_book` for books. For HTML output, use `tufte_html`. These formats can be either specified in the YAML metadata at the beginning of an R Markdown document (see an example below), or passed to the `rmarkdown::render()` function. See @R-rmarkdown for more information about **rmarkdown**.

```yaml
---
title: "An Example Using the Tufte Style"
author: "John Smith"
output:
  tufte::tufte_handout: default
  tufte::tufte_html: default
---
```
```{r, include = FALSE}
require(caret)
require(dplyr)
require(e1071)
require(FMradio)
require(pROC)
require(xgboost)
require(rpart)
require(DMwR2)
require(nnet)
require(ggplot2)
require(ggthemes)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = NA, warning = FALSE
)
```
```{r multifit}
multifit <- function(X,
                     y,
                     model,
                     ctrl = NULL,
                     grid = NULL,
                     seed = 87654, ...) {
  set.seed(seed)
  # Merge X and y into df
  df <- cbind.data.frame(X, y)
  # Train the model
  mdl <- caret::train(df[, 1:ncol(X)], df$y,
    method = model,
    tuneGrid = grid,
    trControl = ctrl,
    metric = "logLoss",
    ...
  )
  # Create a confusion matrix and get performance metrics from caret
  obs <- mdl$pred$obs
  preds <- mdl$pred$pred
  cm <- confusionMatrix(reference = obs, data = preds, mode = "everything")
  # Get the multi-class ROC curves
  ys <- as.numeric(obs) - 1
  yhats <- as.numeric(preds) - 1
  roc <- multiclass.roc(response = ys, predictor = yhats, quiet = T, legacy.axes = T)
  out <- list("cm" = cm, "roc" = roc, "model" = mdl)
  return(out)
}

evaluate <- function(model){
print(model$roc$auc)
print(model$cm)
names(model$roc$rocs) <- c("ADE4-AD", "ADE4-SCDE4", "ADE4-SCD", "AD-SCDE4", "AD-SCD", "SCDE4-SCD")
ggroc(model$roc$rocs, legacy.axes=T) + scale_color_tableau() + theme_tufte()  + guides(color = guide_legend(title = "ROC curves")) 
}
```

### Preparation
```{r multi_preparation}
load("data/data.Rdata")
X <- scale(df[, 1:230])
y <- df$target

Xclin = cbind.data.frame(X, clin_dummy)


ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 100,
  savePredictions = "final",
  classProbs = T,
  summaryFunction = multiClassSummary,
  selectionFunction = best,
  search = "random",
  sampling = "smote"
)
```

### Fitting a benchmark model: clinical characteristics only

```{r mlogit}
benchmark <- multifit(
  X = clin_dummy,
  y = y,
  model = "multinom",
  ctrl = ctrl,
  grid = expand.grid(decay = 0),
  trace = F
)

evaluate(benchmark)
```

### Regularized Multinomial Regression adding 230 metabolites

```{r}
mlr <- multifit(
  X = Xclin,
  y = y,
  model = "multinom",
  ctrl = ctrl,
  grid = expand.grid(decay = 10),
  trace = F
)

evaluate(mlr)
```

### Projection to Latent Factors

```{r get_thomson}
project <- function(X, m, seed) {
  set.seed(seed)
  X <- scale(as.matrix(df[,1:230]))
  cov <- cor(X)

  # Find redundant features
  filter <- RF(cov)

  # Filter out redundant features
  filtered <- subSet(X, filter)

  # Regularized correlation matrix estimation
  M <- regcor(filtered)

  # Get the regularized correlation matrix of the filtered dataset
  R <- M$optCor

  mlfa <- mlFA(R, m = 6)
  thomson <- facScore(filtered, mlfa$Loadings, mlfa$Uniqueness)
  return(thomson)
}
```

```{r}
thomson <- project(X, 6, seed = 1234)
X <- cbind(clin_dummy, thomson)
```

#### Multinomial Logistic Regression adding 6 factors
```{r }
mlrf <- multifit(
  X = X,
  y = y,
  model = "multinom",
  ctrl = ctrl,
  trace = FALSE,
  grid = expand.grid(decay = 0)
)

evaluate(mlrf)
```

#### Decision Tree

```{r multi_tree}
tree <- multifit(
  X = X,
  y = y,
  model = "rpart2",
  ctrl = ctrl,
  grid = expand.grid(maxdepth=3)
)

evaluate(tree)
```

#### XGBoost Forest

```{r }
xgb <- multifit(
  X = X,
  y = y,
  model = "xgbTree",
  ctrl = ctrl,
  grid = xgb.grid
)

evaluate(xgb)
```
## Model Comparison
Observations:


1. Adding serum metabolite information (either the full 230-metabolite matrix or its 6-factor projection) seems to increase the discriminatory power of the models.  


2. Fitting 6 ML-estimated factors obtained by the FMradio package (cummulatively explaining 30% of variace) yields increased classification performance, serving as a valuable dimension reduction technique for high-dimensional data.


3. Looking at the confusion matrix and individual ROC curves, all models were able to discriminate better among certain classes (AD+E4/SCD+E4, AD+E4/SCD, AD-E4/SCD+E4 and AD-E4/SCD-E4) compared to others (AD+E4/AD-E4 and SCD+E4/SCD-E4).

```{r}
aucs <- c(MLR0 = benchmark$roc$auc, MLR1 = mlr$roc$auc, MLR2 = mlrf$roc$auc, TREE = tree$roc$auc, XGB = xgb$roc$auc)
knitr::kable(t(aucs))
```